---
title: "Exploratory Data Analysis"
author: "Fernando de Oliveira Junior"
date: "3/15/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load required packages
library(tidyverse)
library(spotifyr)
library(magrittr)
library(lubridate)
library(corrplot)
library(zoo)
library(psych)

setwd("~/Desktop/TCC")

# read databases
all_songs <- read_csv('2.Datasets/features.csv')
most_popular <- read_csv('2.Datasets/most_popular.csv')

# encode factors features
all_songs$time_signature %<>% as.factor()
all_songs$mode %<>% as.factor()
all_songs$key %<>% as.factor()

most_popular$time_signature %<>% as.factor()
most_popular$mode %<>% as.factor()
most_popular$key %<>% as.factor()

# encode date features
all_songs$track.album.release_date %<>% as.Date()

most_popular$date %<>% as.Date()
most_popular$track.album.release_date %<>% as.Date()

most_popular$ano_mes <- as.yearmon(most_popular$date)
most_popular$ano_mes %<>% factor(ordered = TRUE)

# select unique most popular songs
songs_pop <- most_popular %>%
  select(-c(date, mes, ano, ano_mes)) %>% 
  unique()

# select unique most popular songs monthly
monthly_pop <- most_popular %>%
  select(-c(date)) %>% 
  unique()

# flag each song as pop or not (since Jan 2017)
songs_pop$is_hit <- 1
all_songs %<>%
  left_join(songs_pop[, c(1, 19)], by = "id")

all_songs %<>% mutate(is_hit = coalesce(is_hit, 0))

# asjust the lables of the flags
all_songs$is_hit <- case_when(all_songs$is_hit == 1 ~ "Hit",
                         TRUE ~ "No Hit")

all_songs$is_hit %<>% as.factor()

# create monthly averages of features for most popular songs
ano_mes_pop <- monthly_pop %>% 
  group_by(ano_mes) %>% 
  summarise(popularity = mean(track.popularity),
            danceability = mean(danceability),
            energy = mean(energy),
            loudness = mean(loudness),
            speechiness = mean(speechiness),
            acousticness = mean(acousticness),
            instrumentalness = mean(instrumentalness),
            liveness = mean(liveness),
            valence = mean(valence),
            tempo = mean(tempo),
            duration_ms = mean(duration_ms))
```

### We are going to perform a EDA on Spotify data 

##### First of all, let's see our databases:

##### 1. `All songs`: have features for all `r nrow(all_songs) ` songs generated in  `2020-03-14`.

```{r }
summary(all_songs)
```

##### As we can see there's only NA's in `track.album.release_date`

##### 2. `Most popular`: have features for the 200 most popular songs in Brazil each day since Jan 2017
```{r }
summary(most_popular)
```

---

### Now we are going to state some questions to guide our EDA:
### 1. How is the popularity index calculated?

> The value will be between 0 and 100, with 100 being the most popular. The popularity is calculated by algorithm and is based, in the most part, on the total number of plays the track has had and how recent those plays are.

---

### 2. How is the popularity distributed 

> it has a lot of songs with realy small popularity index and, for other songs, it follows a normal distribuition with a mean of 36.2

```{r echo=FALSE, warning=FALSE, message=FALSE}
all_songs %>%
  ggplot()+
  geom_density(aes(x = track.popularity, y = ..count..), 
                 fill = "#2980b9")+
  theme_minimal()+ 
  geom_vline(aes(xintercept = mean(all_songs$track.popularity)))+
  annotate(geom = "text", 
           x = mean(all_songs$track.popularity)+13, 
           y = 1600, 
           label = paste0("Média: ", round(mean(all_songs$track.popularity), 3)))+
  labs(title = "Distribution of Song Popularity Index",
       subtitle = "All songs",
       x = "Popularity Index",
       y = "")
```

---

### 3. How is the popularity distributed for the songs that were on the 200 charts at any moment since Jan 2017?

> it has a lot of songs with realy small popularity index and, for other songs, it follows a normal distribuition with a mean of 49.5

```{r echo=FALSE, warning=FALSE, message=FALSE}
songs_pop %>%
  ggplot()+
  geom_histogram(aes(x = track.popularity, y = ..count..), 
                 fill = "#2980b9")+
  theme_minimal()+ 
  geom_vline(aes(xintercept = mean(songs_pop$track.popularity)))+
  annotate(geom = "text", 
           x = mean(songs_pop$track.popularity)+13, 
           y = 260, 
           label = paste0("Média: ", round(mean(songs_pop$track.popularity), 3)))+
  labs(title = "Distribution of Song Popularity Index",
       subtitle = "Most Pop Songs (since Jan 2017)",
       x = "Popularity Index",
       y = "")
```

---

### 4. Can we compare, visually, the two distributions?

```{r echo=FALSE, warning=FALSE, message=FALSE}
hit_x <- round(mean(all_songs$track.popularity[all_songs$is_hit == "Hit"]), 3)
no_hit_x <- round(mean(all_songs$track.popularity[all_songs$is_hit != "Hit"]), 3)

all_songs %>%
  ggplot()+
  geom_density(aes(x = track.popularity, 
                   fill = is_hit, 
                   alpha = 0.2,
                   color = is_hit))+
  theme_minimal()+ 
  geom_vline(aes(xintercept = hit_x,),
             color = "#267DB3")+
  annotate(geom = "text", 
           x = hit_x+15, 
           y = 0.03, 
           label = paste0("Média Hits: ", hit_x))+
  geom_vline(aes(xintercept = no_hit_x,),
             color = "#ED6647")+
  annotate(geom = "text", 
           x = 15, 
           y = 0.03, 
           label = paste0("Média No Hits: ", no_hit_x))+
  labs(title = "Distribution of Song Popularity Index",
       subtitle = "Hit Songs vs No Hit Songs",
       x = "Popularity Index",
       y = "")+
  scale_fill_manual(name = "is_hit", values = c("#267DB3", "#ED6647"))+
  scale_color_manual(name = "is_hit", values = c("#267DB3", "#ED6647"))
```

---

### 5. Is there any correlations between song features and popularity index?

> There is no correlation between the features of the songs and popularity, which makes sense, since popularity index is based on changeable parameters

### 5.1 For all songs
```{r echo=FALSE, warning=FALSE, message=FALSE}
corrplot(cor(all_songs[, sapply(all_songs, is.numeric)]), method = "number")
```


### 5.2 For the most popular songs
```{r echo=FALSE, warning=FALSE, message=FALSE}
corrplot(cor(most_popular[, sapply(most_popular, is.numeric)]), method = "number")
```

---

### 6. From 5, do the evolution of features for the most popular songs correlates with popularity over time? It would make sense, if the taste of people is changing over time

> From 6.1 we can say that the popular songs from the past don't, necesseraly have higher popularity today. Acctualy, what we observe is that it's lower than those from the songs that are more popular nowadays, which make sense by the way the index is calculated

> From 6.2 we can observe that the evolution of the features from the most popular song in each month is highly correlated with the popularity index. Threrefore, it can be used in order to create the model to infer the index.

### 6.1 How does popularity distribuition evolve for those songs?
```{r echo=FALSE, warning=FALSE, message=FALSE}
monthly_pop %>%
  ggplot()+
  geom_boxplot(aes(x = ano_mes, y = track.popularity))+
  theme_minimal()+
  scale_x_discrete(breaks = levels(monthly_pop$ano_mes)[c(T, rep(F, 3))])

ano_mes_pop %>%
  ggplot()+
  geom_point(aes(x = ano_mes, y = popularity))+
  geom_smooth(aes(x = seq(1:nrow(ano_mes_pop)), 
                  y = popularity),
              method = lm)+
  labs(title = "Evolution of Mean Popularity of The Most Popular Songs Over Time",
       subtitle = "Linear Regression Tendency",
       x = "Ano Mes",
       y = "")+
  theme_minimal()+
  scale_x_discrete(breaks = levels(ano_mes_pop$ano_mes)[c(T, rep(F, 3))])
```

### 6.2 How is the correlation between the evolution of popularity index and the most popular songs?
```{r echo=FALSE, warning=FALSE, message=FALSE}
corrplot(cor(ano_mes_pop[, sapply(ano_mes_pop, is.numeric)]), method = "number")
```

---

> From 7 we can observe that, on average, each month has 307 songs at the top 200 raking.  

### 7. How many popular (Top 200) songs are there each month?
```{r echo=FALSE, warning=FALSE, message=FALSE}
qnt_monthly_pop <- monthly_pop %>%
  group_by(ano_mes) %>%
  summarise(count = n()) 

qnt_monthly_pop %>%
  ggplot()+
  geom_density(aes(x = count, y = ..count..), 
           fill = "#2980b9")+
  theme_minimal()+
  geom_vline(aes(xintercept = mean(qnt_monthly_pop$count)))+
  annotate(geom = "text",
           x = mean(qnt_monthly_pop$count)+13,
           y = 1,
           label = paste0("Média: ", round(mean(qnt_monthly_pop$count), 3)))+
  labs(title = "Distribution of Songs in Each Month",
       subtitle = "Most Pop Songs (since Jan 2017)",
       x = "Count",
       y = "")
```

---

> From 8.1, there are some outliers and the distribuition is assimetrical for the amount of days a song remains at the top 200 ranking

### 8.1. How many days a song remain at the top 200?
```{r echo=FALSE, warning=FALSE, message=FALSE}
qnt_days_pop <- most_popular %>%
  group_by(id) %>%
  summarise(count = n()) 

qnt_days_pop %>%
  ggplot()+
  geom_density(aes(x = count, y = ..count..), 
               fill = "#2980b9")+
  theme_minimal()+
  geom_vline(aes(xintercept = mean(qnt_days_pop$count)))+
  annotate(geom = "text",
           x = mean(qnt_days_pop$count)+13,
           y = 40,
           label = paste0("Média: ", round(mean(qnt_days_pop$count), 3)))+
  labs(title = "Distribution of Days a Song Remains at the Top 200",
       subtitle = "Most Pop Songs (since Jan 2017)",
       x = "Count",
       y = "")
```

> From 8.2, we are going to drop observations above the percentil 97% (387 days)

### 8.2. Analysing distribuition in order to remove outliers
```{r echo=FALSE, warning=FALSE, message=FALSE}
qnt_days_pop$count %>% 
  quantile(probs = c(seq(0, 0.75, 0.25), seq(0.9, 1, 0.01)))

qnt_days_pop %<>%
  filter(count < 387) 
```

> From 8.3, we can observe that the distribuition is very assimetrical. There are lots of songs with few days in top 200 and a few songs with a lot of days in the top 200 ranking

### 8.3. Analysing the distribuition without outliers
```{r echo=FALSE, warning=FALSE, message=FALSE}
qnt_days_pop %>%
  ggplot()+
  geom_density(aes(x = count, y = ..count..), 
               fill = "#2980b9")+
  theme_minimal()+
  geom_vline(aes(xintercept = mean(qnt_days_pop$count)),
             color = '#8e44ad')+
  geom_vline(aes(xintercept = median(qnt_days_pop$count)),
             color = '#27ae60')+
  annotate(geom = "text",
           x = mean(qnt_days_pop$count)+23,
           y = 40,
           label = paste0("Média: ", round(mean(qnt_days_pop$count), 3)),
           color = '#8e44ad')+
  annotate(geom = "text",
           x = median(qnt_days_pop$count)+23,
           y = 30,
           label = paste0("Mediana: ", round(median(qnt_days_pop$count), 3)),
           color = '#27ae60')+
  labs(title = "Distribution of Days a Song Remains at the Top 200",
       subtitle = "Most Pop Songs (since Jan 2017)",
       x = "Count",
       y = "")
```

---

> From 9, we can see that there are, on average, 132 songs remaning popular in each month, so we could define our monthly songs to be clustered as the songs that remain on top.

### 9. How many songs remain at the top 200 each month?
```{r echo=FALSE, warning=FALSE, message=FALSE}
most_popular$days_in_month <- days_in_month(most_popular$mes)

songs_remaining <- most_popular %>%
  group_by(id, ano_mes, days_in_month) %>%
  summarise(days = n()) %>%
  filter(days_in_month == days)

monthly_remaining <- songs_remaining %>%
  group_by(ano_mes) %>%
  summarise(songs = n())

monthly_remaining %>%
  select(songs) %>%
  summary()
```

---

> From 10, we can see that the median release date for all songs is around 2017 while the songs that were popular (top 200) has the median value around 2018. It's important to address this discussion if we are going to select only songs that were released after 2017

### 10. When were songs released, considering both the populars and all other songs?

### 10.1 All Songs
```{r echo=FALSE, warning=FALSE, message=FALSE}
all_songs %>% 
  select(track.album.release_date) %>% 
  summary()
```

### 10.2 Popular Songs
```{r echo=FALSE, warning=FALSE, message=FALSE}
songs_pop %>% 
  select(track.album.release_date) %>% 
  summary()
```